Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2020-12-21 GWLDCI.tex
some typos, particularly in displays

\begin{dfn}\label{def:edges}
The $i$th edge of $W$ is the arc that lies between the vertices $i$ and $i+1$.
\end{dfn}

@@

\begin{eg}\label{eg of WLD} Let $V = [8]$ and $\cP = \{ (1,4), (2, 4), (5, 8)\}$. Then $W = (\cP,[8])$ is the Wilson loop diagram
\bas \begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
	\drawWLD{8}{1.5}
	\drawnumbers
	\drawprop{1}{0}{4}{0}
	\drawprop{2}{0}{4}{-1}
\drawprop{5}{0}{8}{0}
		\end{scope}
	\end{tikzpicture}  \eas 
\end{eg}

@@

\begin{dfn}\label{admisdfn}
A Wilson loop diagram $W = (\cP,V)$ is {\em admissible} if it satisfies the following three conditions: \begin{enumerate}
\item $|V| \geq |\cP| + 4.$
\item For any non-empty set of propagators, $Q \subseteq \cP$, $Q \neq \emptyset$, one has $|V(Q)| \geq |Q| + 3$.
\item There does not exist a pair of propagators $(i_p,j_p),(i_q,j_q) \in \cP$ such that $i_p < i_q < j_p <j_q$.
\end{enumerate}
A Wilson loop diagram is called {\em weakly admissible} if the second and third conditions hold.
 \end{dfn}

@@

\begin{dfn}\label{equivdfn} 
Let $W = (\cP,[n])$ and $W' = (\cP',[n])$ be two weakly admissible subdiagrams. Let $\sim$ denote the transitive closure of the following relation: $W$ and $W'$ are related if
\begin{enumerate}
\item There exist exact subdiagrams $(P, V(P)) \subseteq W$ and $(P', V(P')) \subseteq W'$ with the same number of propagators and supported on the same vertices, i.e. $|P| = |P'|$ and  $V(P) =  V(P')$.
\item The complementary subdiagrams (i.e. all other propagators and vertices) are identical, i.e. $(\cP \setminus P, [n]) = (\cP' \setminus P', [n])$.
\end{enumerate}
The relation $\sim$ is an equivalence relation on the set of all weakly admissible Wilson loop diagrams with vertex set $[n]$.
\end{dfn}

@@

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,[n])$ be a weakly admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle in $M(W)$, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support disjoint sets of propagators and in such a way that we obtain a partition of the propagator set, i.e. if we have \bas [n] = \sqcup_i V(P_i) \ \textrm{ such that } \ \sqcup P_i = \cP; \   \text{ with }V(P_i) \cap V(P_j) = \emptyset \text{ and } P_i \cap P_j = \emptyset \quad \forall i \neq j\;, \eas then the matroid $M(W)$ is separable, i.e. \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\item The set $F(P)$ from Definition~\ref{VPropdfn} is a flat of $M( W)$, which we call the \emph{propagator flat} of $P$.
\item If $F$ is a cyclic flat of $M(W)$, then it is a propagator flat.
\end{enumerate}
\end{lem}
\begin{proof}
The first part of item 1 is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. Item 2 is \cite[Lemma 3.27]{wilsonloop}. Item 3 is a direct consequence of \cite[Theorem 3.20]{wilsonloop} and the fact that $F(P_1)^c = V(P_1^c)$, and item 5 is given in \cite[Lemma 3.28]{wilsonloop}.

@@

For a weakly admissible Wilson loop diagram $W$, every triangulated piece $t$ of $\tau(W)$ corresponds to a subdiagram of $W$ obtained by taking the set of propagators $P$ corresponding to edges of $t$ and then taking the subdiagram $(P, V(P))$.  Conversely, given a subdiagram $(P, V(P))$ of $W$ we can obtain a subgraph of $\tau(W)$, called $t$, as follows:
\begin{itemize}
\item The vertex set of the subgraph $t$ is
  \begin{itemize}
  \item the vertices of $\tau(W)$ corresponding to edges of $W$ whose end points are elements of $V(P)$ which are cyclically consecutive in $[n]$. (This is significantly easier to interpret in picture form: see Example~\ref{eg:subtrees} below.)
  \end{itemize}
\item The edge set of the subgraph is
  \begin{itemize}
  \item the edges of $\tau(W)$ corresponding to propagators of $P$,
  \item the outer edges of $\tau(W)$ for which both their end points are in the vertex set.
  \end{itemize}
\end{itemize}

@@

Since Corollary \ref{uniqueproppartitioncor} allows us to decompose any admissible Wilson loop diagram into a collection of maximal exact subdiagrams, we are motivated to examine the matroid properties of exact subdiagrams more closely. In this section, we prove two main results:
\begin{enumerate}
\item The matroid associated to an exact subdiagram of $W$ can be written as a contraction of the matroid $M(W)$ by the complementary propagator flat (Theorem \ref{exact diagrams contractions}). 
\item The matroid associated to an exact subdiagram is uniform (Theorem \ref{exactuniformthm}).
\end{enumerate}

@@

\begin{lem} \label{lem decompose flat}Let $F$ be a flat in $M(W)$, and let $C \subseteq F$ be the union of all circuits contained in $F$. Then the following are true:
\begin{enumerate}
\item $C = F(\Prop (C))$, i.e. $C$ is a propagator flat.
\item $F \setminus C$ is an independent set. Furthermore, $F\setminus C$ is an independent flat if and only if $F(\emptyset) = \emptyset$, that is, $W$ has no non-supporting vertices.  
\end{enumerate}
\end{lem}

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $F(P^c)$ is a dependent flat in $W$.
\item If $(P,V(P))$ is a maximal exact subdiagram of $W$, then $F(P^c)$ is a cyclic flat in $W$.
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate} 
\end{lem}

@@

For the if direction, it suffices to prove the result for two equivalent diagrams that differ in exactly one exact subdiagram. Let $W$ and $W'$ be equivalent Wilson loop diagrams which can be decomposed as
\[W = (P \sqcup R, [n]), \qquad W' = (P \sqcup R', [n]),\]
where $(R,V(R))$ and $(R',V(R'))$ are different maximal exact subdiagrams of $W$ and $W'$ respectively, and $V(R) = V(R')$.

@@

\begin{cor}\label{number of equiv diagrams}
  Let $W$ be a weakly admissible Wilson loop diagram where the sizes of the supports of the nontrivial maximal exact subdiagrams are $n_1, n_2, \ldots, n_j$.  Then the number of weakly admissible Wilson loop diagrams equivalent to $W$ (including $W$ itself) is
  \[
  \prod_{i=1}^{j} \frac{1}{n_i-1}\binom{2(n_i-2)}{n_i-2}
  \]
\end{cor}

@@

\begin{lem}\label{lem good quads}
Let $D_1$ and $D_2$ be two dissections of a convex $n$-gon defined by the vertices $x_1, \ldots, x_n$.  Suppose that $V_{>3}(D_1) = V_{>3}(D_2)$, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ satisfying the following conditions:\begin{enumerate} \item In either $D_1$ or $D_2$ all four vertices are on the same non-triangle face. \item In the other dissection, there is a diagonal of the dissection such that exactly one of $\{x_i, x_j, x_k, x_l\}$ lies on one side of the diagonal (so the others are either end points of this diagonal or strictly on the other side). \end{enumerate}
\end{lem}

>>>>

\begin{dfn}\label{def:edges}
The $i^{th}$ edge of $W$ is the arc that lies between the vertices $i$ and $i+1$.
\end{dfn}

@@

\begin{eg}\label{eg of WLD} Let $V = [8]$ and $\cP = \{ (1,4), (2, 4), (5, 8)\}$. Then $W = (\cP,[8])$ is the Wilson loop diagram
\bas \begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
	\drawWLD{8}{1.5}
	\drawnumbers
	\drawprop{1}{0}{4}{0}
	\drawprop{2}{0}{4}{-1}
\drawprop{5}{0}{8}{0}
		\end{scope}
	\end{tikzpicture} \; . \eas 
\end{eg}

@@

\begin{dfn}\label{admisdfn}
A Wilson loop diagram $W = (\cP,V)$ is {\em admissible} if it satisfies the following three conditions: \begin{enumerate}
\item $|V| \geq |\cP| + 4,$
\item for any non-empty set of propagators, $Q \subseteq \cP$, $Q \neq \emptyset$, one has $|V(Q)| \geq |Q| + 3$.\,
\item there does not exist a pair of propagators $(i_p,j_p),(i_q,j_q) \in \cP$ such that $i_p < i_q < j_p <j_q$.
\end{enumerate}
A Wilson loop diagram is called {\em weakly admissible} if the second and third conditions hold.
 \end{dfn}

@@

\begin{dfn}\label{equivdfn} 
Let $W = (\cP,[n])$ and $W' = (\cP',[n])$ be two weakly admissible subdiagrams. Let $\sim$ denote the transitive closure of the following relation: $W$ and $W'$ are related if
\begin{enumerate}
\item there exist exact subdiagrams $(P, V(P)) \subseteq W$ and $(P', V(P')) \subseteq W'$ with the same number of propagators and supported on the same vertices, i.e. $|P| = |P'|$ and  $V(P) =  V(P')$.
\item the complementary subdiagrams (i.e. all other propagators and vertices) are identical, i.e. $(\cP \setminus P, [n]) = (\cP' \setminus P', [n])$.
\end{enumerate}
The relation $\sim$ is an equivalence relation on the set of all weakly admissible Wilson loop diagrams with vertex set $[n]$.
\end{dfn}

@@

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,[n])$ be a weakly admissible Wilson loop diagram. The following are true.
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle in $M(W)$, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support disjoint sets of propagators and in such a way that we obtain a partition of the propagator set, i.e. if we have \bas [n] = \sqcup_i V(P_i) \ \textrm{ such that } \ \sqcup P_i = \cP; \   \text{ with }V(P_i) \cap V(P_j) = \emptyset \text{ and } P_i \cap P_j = \emptyset \quad \forall i \neq j\;, \eas then the matroid $M(W)$ is separable, i.e. \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\item The set $F(P)$ from Definition~\ref{VPropdfn} is a flat of $M( W)$, which we call the \emph{propagator flat} of $P$.
\item If $F$ is a cyclic flat of $M(W)$, then it is a propagator flat.
\end{enumerate}
\end{lem}
\begin{proof}
The first part of item 1 is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. Item 2 is \cite[Lemma 3.27]{wilsonloop}. Item 3 is a direct consequence of \cite[Theorem 3.20]{wilsonloop} and the fact that $F(P_1)^c = V(P_1^c)$, and item 5 is given in \cite[Lemma 3.28]{wilsonloop}.

@@

For a weakly admissible Wilson loop diagram $W$, every triangulated piece $t$ of $\tau(W)$ corresponds to a subdiagram of $W$ obtained by taking the set of propagators $P$ corresponding to edges of $t$ and then taking the subdiagram $(P, V(P))$.  Conversely, given a subdiagram $(P, V(P))$ of $W$ we can obtain a subgraph of $\tau(W)$, called $t$, following the algorithm below.
\begin{itemize}
\item The vertex set of the subgraph $t$ is
  \begin{itemize}
  \item the vertices of $\tau(W)$ corresponding to edges of $W$ whose end points are elements of $V(P)$ which are cyclically consecutive in $[n]$. (This is significantly easier to interpret in picture form: see Example~\ref{eg:subtrees} below.)
  \end{itemize}
\item The edge set of the subgraph is
  \begin{itemize}
  \item the edges of $\tau(W)$ corresponding to propagators of $P$,
  \item the outer edges of $\tau(W)$ for which both their end points are in the vertex set.
  \end{itemize}
\end{itemize}

@@

Since Corollary \ref{uniqueproppartitioncor} allows us to decompose any admissible Wilson loop diagram into a collection of maximal exact subdiagrams, we are motivated to examine the matroid properties of exact subdiagrams more closely. In this section, we prove two main results.
\begin{enumerate}
\item the matroid associated to an exact subdiagram of $W$ can be written as a contraction of the matroid $M(W)$ by the complementary propagator flat (Theorem \ref{exact diagrams contractions}),
\item the matroid associated to an exact subdiagram is uniform (Theorem \ref{exactuniformthm}).
\end{enumerate}

@@

\begin{lem} \label{lem decompose flat}Let $F$ be a flat in $M(W)$, and let $C \subseteq F$ be the union of all circuits contained in $F$. Then the following are true:
\begin{enumerate}
\item $C = F(\Prop (C))$, i.e. $C$ is a propagator flat,
\item $F \setminus C$ is an independent set. Furthermore, $F\setminus C$ is an independent flat if and only if $F(\emptyset) = \emptyset$, that is, $W$ has no non-supporting vertices.  
\end{enumerate}
\end{lem}

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram, and $P \subseteq \cP$. Then, \begin{enumerate}
\item if $(P,V(P))$ is an exact subdiagram of $W$, then $F(P^c)$ is a dependent flat in $W$,
\item if $(P,V(P))$ is a maximal exact subdiagram of $W$, then $F(P^c)$ is a cyclic flat in $W$,
\item if $(P,V(P))$ is an exact subdiagram of $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate} 
\end{lem}

@@

For the if direction, it suffices to prove the result for two equivalent diagrams that differ in exactly one exact subdiagram. Let $W$ and $W'$ be equivalent Wilson loop diagrams which can be decomposed as
\[W = (P \sqcup R, [n]) \qquad \textrm{and} \qquad W' = (P \sqcup R', [n]),\]
where $(R,V(R))$ and $(R',V(R'))$ are different maximal exact subdiagrams of $W$ and $W'$ respectively, and $V(R) = V(R')$.

@@

\begin{cor}\label{number of equiv diagrams}
  Let $W$ be a weakly admissible Wilson loop diagram where the sizes of the supports of the nontrivial maximal exact subdiagrams are $n_1, n_2, \ldots, n_j$.  Then the number of weakly admissible Wilson loop diagrams equivalent to $W$ (including $W$ itself) is
  \[
  \prod_{i=1}^{j} \frac{1}{n_i-1}\binom{2(n_i-2)}{n_i-2} \;.
  \]
\end{cor}

@@

\begin{lem}\label{lem good quads}
Let $D_1$ and $D_2$ be two dissections of a convex $n$-gon defined by the vertices $x_1, \ldots, x_n$.  Suppose that $V_{>3}(D_1) = V_{>3}(D_2)$, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ satisfying the following conditions:\begin{enumerate} \item in either $D_1$ or $D_2$ all four vertices are on the same non-triangle face. \item in the other dissection, there is a diagonal of the dissection such that exactly one of $\{x_i, x_j, x_k, x_l\}$ lies on one side of the diagonal (so the others are either end points of this diagonal or strictly on the other side). \end{enumerate}
\end{lem}
